Unreliable heterogeneity: how measurement error obscures heterogeneity in meta-analyses in psychology

### 2023-06-17 manuscript/manuscript.qmd
dynamic references

---
title: "Unreliable heterogeneity: how measurement error obscures heterogeneity in meta-analyses in psychology"
date: today
date-format: long
author:
  - name: Anton Olsson-Collentine
    url: https://example.com
    orcid: 0000-0000-0000-0000
    affiliations:
      - id: tiu
        name: "Department of Methodology and Statistics, Tilburg School of Social and Behavioral Sciences, Tilburg University, the Netherlands."
    email: anton@olssoncollentine.com
  - name: Marjan Bakker
    affiliations:
      - ref: tiu
  - name: Jelte Wicherts
    affiliations:
      - ref: tiu
format:
  bj-pdf:
    classoption:
      - twocolumn
colorlinks: true
link-citations: true
keep-tex: false
abstract: |
  One of the main purposes of meta-analysis is to estimate heterogeneity of effect sizes. The quality of meta-analytic estimates depends in large part on the quality of evidence in primary studies, yet there exist broad concerns that proper measurement tends to be neglected in much of psychological primary research. Such inattention to measurement can also bias heterogeneity estimates. We performed a Monte-Carlo simulation study to examine the expected bias in heterogeneity estimates due to unreliability in primary studies. Although unreliability in primary studies can both inflate and suppress heterogeneity, our results indicate that under most circumstances unreliability in primary studies leads to a severe underestimate of heterogeneity in meta-analyses. For a typical effect size heterogeneity can be expected to be underestimated by 15-60%, or more if average reliability in primary studies is below 0.6. We observed a positive bias in heterogeneity estimates due to unreliability only under specific and arguably uncommon circumstances of (1) actual zero heterogeneity, particularly when mean effect sizes are large, or (2) combinations of very small true heterogeneity, large variance in primary study reliabilities, large mean effect sizes, and a limited number of primary studies. The severity of the underestimation of heterogeneity could lead to meta-analyses in psychology missing relevant moderators and overgeneralizing outcomes. Unreliability appears to bring truly different effects closer together, thereby obscuring true differences between studies that could be relevant for theory, practice, and future research efforts. Research on concrete guidance to applied meta-analysts is urgently needed, as sophisticated methods for correcting measurement unreliability such as meta-analytic structural equation modeling (MASEM) are only applicable in exceptional cases and corrections based on classical test theory come with caveats and strong assumptions.
license: CC BY
keywords:
  - Research
  - Cool Stuff
citation:
  type: article
  citation-key: citationKeyNumber
  status: draft
cite-method: citeproc
csl: apa.csl
bibliography: references.bib
---

@@

Estimating heterogeneity of effect sizes is usually considered the main purpose of meta-analysis, alongside estimating an average effect size. Heterogeneity of effect sizes (henceforth referred to as heterogeneity) refers to an effect size’s sensitivity to variability in study design features related to the type of (1) sample, (2) setting, (3) treatment variable, and (4) measurement (e.g., Shadish, Cook, & Campbell, 2002). Heterogeneity is an important outcome in meta-analysis for several reasons. First, its presence signals the existence of moderators and so can be seen as an opportunity for theoretical development (Simons et al., 2017). Second, heterogeneity affects the implementation of research by indicating that an intervention may not be equally effective under all conditions or for everyone. Third, the presence of heterogeneity changes the interpretation of the average effect size derived in a meta-analysis from the true population effect size (under homogeneity) to the average of true subpopulation effect sizes (under heterogeneity). As such, estimating heterogeneity with minimal bias is of central importance to meta-analysis (for a more extensive discussion on the importance of heterogeneity, see Olsson Collentine et al., 2020; Simons, 2017).

Meta-analytic estimates depend on the quality of data in primary studies. However, there is concern that researchers in psychology tend to neglect proper measurement (Flake & Fried, 2020) to the extent that psychology can be said to be in a measurement crisis (Kane et al., 2021). For example, Flake et al. (2017) report that although latent constructs are widely examined in personality and social psychology, researchers barely report any of the validity evidence required to ascertain the extent to which implemented scales measure the constructs of interest. Such inattention to measurement can also bias heterogeneity estimates. We will focus here on the reliability of measurement (defined in classical test theory as the ratio of true score variance to measurement error variance; e.g., Lord & Novick, 1968), which is the only  evidence on measurement quality that researchers report consistently (Flake et al., 2017), and which can be corrected for in meta-analysis (Hedges & Olkin, 1985; Hunter & Schmidt; 2015).

Measurement error (imperfect reliability) is present in any empirical effect size estimate and systematically attenuates observed effect sizes compared to true underlying effect sizes (Spearman, 1904). This affects heterogeneity estimates, which from a statistical perspective, estimate the variance between true effect sizes. However, reliability is not necessarily accounted for in heterogeneity estimates, because ‘true’ underlying effect sizes generally refer to two different entities: Either effect sizes free from sampling error, or effect sizes free from both sampling error and measurement error. Whereas all meta-analytic models attempt to correct for sampling error in their estimates, correcting for measurement errors is explicitly modelled (alongside other measurement artifacts) only in the psychometric meta-analysis approach by Hunter & Schmidt (2015). This has consequences for heterogeneity estimates reported in many areas of psychology. Inattention to measurement reliability and how it affects heterogeneity estimates can lead meta-analysts to misinterpret their average effect size estimate, ignore the presence of theoretically or practically relevant moderators or search for (and discover) non-existent moderators, over-generalize outcomes, and implement research interventions inappropriately.

Meta-analyses in psychology rarely correct for unreliability in primary study measurements. Wiernick and Dahlke (2020) reported that among the 71 meta-analytic studies published in the journal Psychological Bulletin between 2016 and 2018, only 6/71 (8%) corrected for unreliability. Similarly, Schmidt (2010) reported that only 19/199 (10%) of the meta-analytic studies published in the same journal between 1978 - 2006 corrected for any measurement artifacts. The exception is the subfield of industrial-organizational studies, where corrections tend to be more common (e.g., Cortina, 2003; Anguinis, 2011).

One reason that few meta-analysts correct for unreliability could be that correcting for unreliability has long been controversial (for arguments spanning the last century, see Table 2 of LeBreton et al., 2017). On the one hand, correcting for unreliability is seen by some as conceptually problematic because it inflates effect size estimates to match a hypothetical and unachievable scenario of perfect measurement (e.g., Seymour, 1988; LeBreton et al., 2017). Systematically correcting for unreliability may also lead to an overestimation of effect sizes, because reliability estimates are lower bound estimates of the true reliability (i.e., a reliability estimate of 0.8 implies the reliability is between 0.8 - 1). Even worse, the most popular estimate of reliability in psychology (Cronbach’s Alpha; Flake et al., 2017), as well as improvements thereof, tend to underestimate this lower bound of reliability (Sijtsma, 2008). Moreover, from a pragmatic meta-analytic perspective, effect sizes appear to often be overestimated in psychology (e.g., Kvarven et al., 2019; Schäfer & Scharz, 2019; Scheel et al., 2020) due to selective reporting based on the significance of outcomes (Simmons et al., 2011) and publication bias (Thornton & Lee, 2000), so correcting for unreliability may inflate meta-analytic estimates even further and lead to estimates further from their true values rather than closer. On the other hand, researchers in psychology are typically interested in latent constructs rather than observed measures. As such, neglecting measurement errors means that computed estimates do not correspond to the entity of interest. From this perspective, correcting for unreliability is desirable even if doing so is challenging (e.g., Oswald et al., 2015; Hunter & Schmidt, 2015), although treating disattenuated effect sizes as directly comparable to latent scores requires strong assumptions (Borsboom & Mellenbergh, 2002). In the end, whether correcting for unreliability or not, authors and consumers of meta-analyses need to be aware of how unreliability of measurement affects heterogeneity estimates. Hunter & Schmidt (2015) assert that failure to control for variation in unreliability across studies results in a large overestimate of heterogeneity (p. 159), but little systematic research has been done to study the effects of unreliability on estimating heterogeneity.

@@

In classical test theory (CTT), reliability of measurement ($R_{xx}$) is defined as the proportion true score variance ($\sigma^2_T$) to observed score variance ($\sigma^2_X$) and is presumed to be constant for all participants in a sample $R_{xx} = \sigma^2_T / \sigma^2_X$. We use unreliability of measurement to refer to $1 - R_{xx}$. Imperfect reliability in measurements leads to attenuation of observed effect sizes. For product-moment correlations the observed correlation can be computed as $r_{xy} = \rho_{xy} \times \sqrt{R_{xx'}} \times \sqrt{R_{yy'}}$ where $r_{xy}$ is the observed correlation between the variables $X$ and $Y$, $\rho_{xy}$ is the true correlation (with or without sampling error) and $R_{xx'}$ and $R_{yy'}$ are the measurement reliabilities for $X$ and $Y$. As such, correcting for unreliability is straightforward for product-moment correlations, but can also be done for other effect size types (Wiernick and Dahlke, 2020). A corrected correlation should also have its sampling variance corrected. This is inflated by a factor equal to the square of the reliabilities, such that if $a = \sqrt{R_{xx'}} \times \sqrt{R_{yy'}}$ then the corrected sampling variance $V_\rho = V_{r} / a^2$ (e.g., Borenstein, 2009, p. 343). Reliability is usually assumed to be known (but see Charles, 2005; Raju et al., 1991), Imperfect reliability in primary studies can both increase and suppress true heterogeneity.

@@

Heterogeneity is suppressed in heterogeneity estimates of uncorrected effect sizes because in absolute terms larger effect sizes are more attenuated by a given reliability. That is, although the attenuation formula $r_{xy} = \rho_{xy} \times \sqrt{R_{xx'}} \times \sqrt{R_{yy'}}$ implies that any effect size $\rho_{xy}$ will be attenuated by the same proportion given the same reliabilities $\sqrt{R_{xx'}} \times \sqrt{R_{yy'}}$, this proportion corresponds to a larger absolute value for larger effect sizes. Consequently, in the presence of heterogeneity attenuation will move larger true effect sizes further towards zero than smaller ones, decreasing heterogeneity in observed effect sizes. This suppression can be observed directly by assuming an average effect size $\theta$ with some heterogeneity $\tau^2$, infinite sample size within studies (i.e., no within-study sampling variance), and equal reliability across studies ($R = R_i = R_{xx'} = R_{yy'}$). Any observed effect size $\hat{\theta_i}$ can then be seen as a study’s true effect size $\theta_i$ multiplied by the study reliability $R$. Hence, the distribution of observed effect sizes is the distribution of true effect sizes $N(\theta, \tau^2)$ multiplied by the reliability $R$, such that $R \times N(\theta, \tau^2) = N(R \theta, R^2 \tau^2) = N(R \theta, R \tau)$. Because reliability lies between 0 – 1, under these conditions imperfect reliability implies that both the observed average effect size and the estimated heterogeneity will be less than their true value by a factor $\bar{R}$ (if heterogeneity is expressed as standard deviation rather than variance). This depiction is not entirely correct for Pearson's $r$ as it is bounded at {-1, 1}, but holds approximately if the mean effect size $\theta$ and heterogeneity $\tau^2$ are not too large. Hence, imperfect reliability in studies can be expected to suppress heterogeneity estimates whenever true heterogeneity is present. Consider an average reliability of 0.8 (the median reported reliability in psychology; Flake et al., 2017). Then, $0.8 \times N(\theta, \tau) = N(0.8\theta, 0.8 \tau)$, resulting in 20% less standard deviation among observed effect sizes than in true effect sizes. Table 2 illustrates this relationship for infinite sample size (i.e., no sampling variance) and three levels of reliability across three studies.

@@

It is difficult to predict if the total effect is a negative or positive bias in heterogeneity due to unreliability, and existing literature tends to focus on the inflation of heterogeneity (e.g., Borenstein, 2009, p. 342; Card, 2015, p. 126; Hunter & Schmidt, 2015; Wiernick & Dahlke, 2020). However, given the value of accurate heterogeneity estimates for many research outcomes in psychology and that the suppression of heterogeneity could often be 20% or more, we consider it important for meta-analysts to have insight into the bias in heterogeneity estimates that should be expected due to unreliability in a wider range of settings. The effect of unreliability depends on average effect size, true heterogeneity, mean reliability across studies, the variability in reliabilities, and sampling variance within studies. Hence, here we perform a Monte-Carlo simulation study to explore the expected bias in heterogeneity estimates due to unreliability in primary studies. In our simulation, we consider reliabilities to be known and conforming to the strict assumptions of CTT to gain a better understanding of how measurement error biases heterogeneity estimates, while acknowledging that in meta-analytic practice, heterogeneity estimation is further challenged by selective reporting, sampling error, and estimation of reliabilities, which is particularly challenging when CTT’s strict assumptions are violated.

@@

All code and data for this project are available at (OSF LINK). For our simulations and analyses we used R (REF) and we used the R package 'metafor' (REF) to perform meta-analyses and estimate heterogeneity. For parallel computing we took advantage of the R packages 'parallel' (REF) and 'parabar' (REF).

@@

We focus on correlations as our effect size type in these simulations because correcting them for unreliability is straightforward and because they are the effect size type most corrected in practice. However, unreliability attenuates all types of effect sizes and all can be corrected, either directly (standardized mean differences; Wiernick & Dahlke, 2020) or by first transforming them into correlations. Although effect sizes can generally be transformed into each other (e.g., Borenstein, 2009, p. 46), this should be done with care as it can sometimes affect conclusions (e.g., we found a violation of monotonicity in heterogeneity estimates when converting effect sizes into correlations in a previous project; Olsson-Collentine, 2020, supplement A).

Perhaps the most common way to meta-analyze correlations is to first transform them into Fisher z correlations (Borenstein et al., 2009, p. 41 - 43) to normalize their distribution. The Fisher $z$ transformation is the inverse hyperbolic tangent of the product moment correlation and can be computed as $z = 0.5 \times ln\left(\frac{1+r}{1-r}\right)$ and its variance approximated as $V_z = 1 / (n -3)$. There is controversy in the use of the Fisher z transformation for meta-analysis (e.g., Hunter & Schmidt, 2015, p. 220 - 223), largely related to whether Pearson’s r or Fisher’s z leads to less bias in average correlations (Schulze, 2007). However, estimating heterogeneity does not seem to have been the focus of most of the literature’s discussion of the appropriateness of the Fisher z transformation (Field, 2005; Hafdahl & Williams, 2009, Hunter & Schmidt, 2015). Hall and Brannick (2002) do report heterogeneity estimates but focus on the coverage of prediction intervals (‘credibility intervals’ in Hunter & Schmidt terminology) which combine the heterogeneity estimate and point estimate, and Brannick et al. (2019) examined results only after corrections for attenuation. Given both this inconclusive evidence and the common use of the Fisher z transformation in practice, we use it as one of the effect size types in our simulations.

We also use Pearson’s r. Pearson’s r has the advantage that it creates heterogeneity estimates on the same scale that are more interpretable than heterogeneity estimates on the Fisher z scale. Although Fisher z correlations can be back-transformed into Pearson’s r for interpretation, its heterogeneity cannot. A workaround is to compute a prediction interval and convert this interval onto the product-moment scale (Hedges & Vevea, 1998). The width of the interval can then be used as an estimate of the heterogeneity. We prefer to compute comparable heterogeneity values for Pearson’s r and Fisher’s z (see the methods section ‘Parameter values’) and report the bias in heterogeneity estimates, as we believe this is more interpretable for applied meta-analysts.

The disadvantage to using Pearson’s r when estimating heterogeneity are that (1) the effect size is bounded to {-1, 1} and (2) its sampling variance $\sigma^2_r = \frac{(1-\rho^2)^2}{n - 1}$ depends on the true effect size $\rho$. The bounded nature of the Pearson correlation means that large levels of heterogeneity create substantial truncation of the distribution of effect sizes if the average effect size is also large. That the sampling variance covaries with effect size leads to a small positive bias in heterogeneity estimates (supplement A). This can be solved by replacing $\rho$ with the average correlation across studies $\bar{r}$ (Hunter & Schmidt, 2015), or using the Fisher $z$ transformation. Fisher z correlations have the additional advantage that they are not bound to {-1, 1}. A disadvantage to using Fisher z correlations, in addition to interpretability of heterogeneity estimates, is that they require back-transforming into product-moment correlations before adding measurement error, followed by another transformation to Fisher z for meta-analysis. Especially for large Fisher z values (which are common when both average effect size and heterogeneity are large) this transformation process may introduce inaccuracies in the estimates. Given the common use of both effect size types and their mix of advantages and disadvantages we report results for both Fisher’s z and Pearson’s r.

## Meta-analytic model

We apply two meta-analytic models to ensure that our results are not model-dependent: the Hedges and Vevea (1998) random-effects model and the Hunter and Schmidt (2015) ‘bare-bones’ random-effects model. The Hedges and Vevea (HV) model uses inverse variance weight computed as $w_i = 1 / (V_i + T^2)$ (e.g., Borenstein et al., 2010) where $V_i$ is the sampling variance of study $i$ and $T^2$ is the estimated between-studies variance. The Hunter and Schmidt (HS) model weighs studies by sample size $w_i = N_i$. The differences between these weights are relatively minor (Borenstein et al., 2010; Hunter and Schmidt, 2015, p. 423-426). Although the HV model is typically applied to Fisher z transformed correlations (e.g., Borenstein et al., p. 41-43), we apply it to both transformed and untransformed correlations to enable a more direct comparison with the HS model and facilitate interpretation. The HS model typically includes corrections for unreliability or other measurement artifacts. However, we implement it without corrections (‘bare-bones’; Hunter & Schmidt, 2015) because this allows us to focus on the bias in heterogeneity estimates rather than the performance of corrections (e.g., see Zhang, 2022 for how well they perform in terms of mean estimates).

```{r Figure 1, out.width= "85%"}
#| fig.env: figure*
#| fig-pos: 't'
#| fig-cap: "Figure 1. Flowchart summarizing the simulation design. The flowchart shows the procedures followed to meta-anlyze results using the Hedge’s and Vevea model (see ‘Meta-Analytic Model’ header in methods section). The Hunter & Schmidt ‘bare-bones’ meta-analytic model followed the Pearson correlations procedure, but used sample size weights and replaced $r_i$ in the sampling variance by r for meta-anlysis."

>>>>

---
title: "Unreliable heterogeneity: how measurement error obscures heterogeneity in meta-analyses in psychology"
date: today
date-format: long
author:
  - name: Anton Olsson-Collentine
    url: https://example.com
    orcid: 0000-0000-0000-0000
    affiliations:
      - id: tiu
        name: "Department of Methodology and Statistics, Tilburg School of Social and Behavioral Sciences, Tilburg University, the Netherlands."
    email: anton@olssoncollentine.com
  - name: Marjan Bakker
    affiliations:
      - ref: tiu
  - name: Jelte Wicherts
    affiliations:
      - ref: tiu
format:
  bj-pdf:
    classoption:
      - twocolumn
colorlinks: true
link-citations: true
keep-tex: false
abstract: |
  One of the main purposes of meta-analysis is to estimate heterogeneity of effect sizes. The quality of meta-analytic estimates depends in large part on the quality of evidence in primary studies, yet there exist broad concerns that proper measurement tends to be neglected in much of psychological primary research. Such inattention to measurement can also bias heterogeneity estimates. We performed a Monte-Carlo simulation study to examine the expected bias in heterogeneity estimates due to unreliability in primary studies. Although unreliability in primary studies can both inflate and suppress heterogeneity, our results indicate that under most circumstances unreliability in primary studies leads to a severe underestimate of heterogeneity in meta-analyses. For a typical effect size heterogeneity can be expected to be underestimated by 15-60%, or more if average reliability in primary studies is below 0.6. We observed a positive bias in heterogeneity estimates due to unreliability only under specific and arguably uncommon circumstances of (1) actual zero heterogeneity, particularly when mean effect sizes are large, or (2) combinations of very small true heterogeneity, large variance in primary study reliabilities, large mean effect sizes, and a limited number of primary studies. The severity of the underestimation of heterogeneity could lead to meta-analyses in psychology missing relevant moderators and overgeneralizing outcomes. Unreliability appears to bring truly different effects closer together, thereby obscuring true differences between studies that could be relevant for theory, practice, and future research efforts. Research on concrete guidance to applied meta-analysts is urgently needed, as sophisticated methods for correcting measurement unreliability such as meta-analytic structural equation modeling (MASEM) are only applicable in exceptional cases and corrections based on classical test theory come with caveats and strong assumptions.
license: CC BY
keywords:
  - Research
  - Cool Stuff
citation:
  type: article
  citation-key: citationKeyNumber
  status: draft
cite-method: citeproc
csl: apa.csl
bibliography: [unreliable-heterogeneity.bib]
---

@@

Estimating heterogeneity of effect sizes is usually considered the main purpose of meta-analysis, alongside estimating an average effect size. Heterogeneity of effect sizes (henceforth referred to as heterogeneity) refers to an effect size’s sensitivity to variability in study design features related to the type of (1) sample, (2) setting, (3) treatment variable, and (4) measurement [e.g., @cook2002]. Heterogeneity is an important outcome in meta-analysis for several reasons. First, its presence signals the existence of moderators and so can be seen as an opportunity for theoretical development [@10.1177/1745691617708630]. Second, heterogeneity affects the implementation of research by indicating that an intervention may not be equally effective under all conditions or for everyone. Third, the presence of heterogeneity changes the interpretation of the average effect size derived in a meta-analysis from the true population effect size (under homogeneity) to the average of true subpopulation effect sizes (under heterogeneity). As such, estimating heterogeneity with minimal bias is of central importance to meta-analysis [for a more extensive discussion on the importance of heterogeneity, see @10.1037/bul0000294; @10.1177/1745691617708630].

Meta-analytic estimates depend on the quality of data in primary studies. However, there is concern that researchers in psychology tend to neglect proper measurement [@10.1177/2515245920952393] to the extent that psychology can be said to be in a measurement crisis [@10.3758/s13428-021-01557-x]. For example, @10.1177/1948550617693063 report that although latent constructs are widely examined in personality and social psychology, researchers barely report any of the validity evidence required to ascertain the extent to which implemented scales measure the constructs of interest. Such inattention to measurement can also bias heterogeneity estimates. We will focus here on the reliability of measurement [defined in classical test theory as the ratio of true score variance to measurement error variance; e.g., @lord1968], which is the only  evidence on measurement quality that researchers report consistently [@10.1177/1948550617693063], and which can be corrected for in meta-analysis [@hedges1985; @schmidt2015] .

Measurement error (imperfect reliability) is present in any empirical effect size estimate and systematically attenuates observed effect sizes compared to true underlying effect sizes [@10.2307/1412159]. This affects heterogeneity estimates, which from a statistical perspective, estimate the variance between true effect sizes. However, reliability is not necessarily accounted for in heterogeneity estimates, because ‘true’ underlying effect sizes generally refer to two different entities: Either effect sizes free from sampling error, or effect sizes free from both sampling error and measurement error. Whereas all meta-analytic models attempt to correct for sampling error in their estimates, correcting for measurement errors is explicitly modelled (alongside other measurement artifacts) only in the psychometric meta-analysis approach by @schmidt2015. This has consequences for heterogeneity estimates reported in many areas of psychology. Inattention to measurement reliability and how it affects heterogeneity estimates can lead meta-analysts to misinterpret their average effect size estimate, ignore the presence of theoretically or practically relevant moderators or search for (and discover) non-existent moderators, over-generalize outcomes, and implement research interventions inappropriately.

Meta-analyses in psychology rarely correct for unreliability in primary study measurements. @10.1177/2515245919885611 reported that among the 71 meta-analytic studies published in the journal Psychological Bulletin between 2016 and 2018, only 6/71 (8%) corrected for unreliability. Similarly, @10.1177/1745691610369339 reported that only 19/199 (10%) of the meta-analytic studies published in the same journal between 1978 - 2006 corrected for any measurement artifacts. The exception is the subfield of industrial-organizational studies, where corrections tend to be more common [e.g., @10.1177/0149206310377113; @10.1177/1094428103257358].

One reason that few meta-analysts correct for unreliability could be that correcting for unreliability has long been controversial [for arguments spanning the last century, see Table 2 of @10.1111/iops.12184]. On the one hand, correcting for unreliability is seen by some as conceptually problematic because it inflates effect size estimates to match a hypothetical and unachievable scenario of perfect measurement (e.g., @10.1016/0001-87918890043-7; @10.1111/iops.12184]. Systematically correcting for unreliability may also lead to an overestimation of effect sizes, because reliability estimates are lower bound estimates of the true reliability (i.e., a reliability estimate of 0.8 implies the reliability is between 0.8 - 1). Even worse, the most popular estimate of reliability in psychology [Cronbach’s Alpha; @10.1177/1948550617693063], as well as improvements thereof, tend to underestimate this lower bound of reliability [@10.1007/s11336-008-9101-0]. Moreover, from a pragmatic meta-analytic perspective, effect sizes appear to often be overestimated in psychology [e.g., @10.1038/s41562-019-0787-z; @10.3389/fpsyg.2019.00813; @10.1177/25152459211007467] due to selective reporting based on the significance of outcomes [@10.1177/0956797611417632] and publication bias [@10.1016/S0895-43569900161-4], so correcting for unreliability may inflate meta-analytic estimates even further and lead to estimates further from their true values rather than closer. On the other hand, researchers in psychology are typically interested in latent constructs rather than observed measures. As such, neglecting measurement errors means that computed estimates do not correspond to the entity of interest. From this perspective, correcting for unreliability is desirable even if doing so is challenging [e.g., @10.1017/iop.2015.17; @schmidt2015], although treating disattenuated effect sizes as directly comparable to latent scores requires strong assumptions [@10.1016/S0160-28960200082-X]. In the end, whether correcting for unreliability or not, authors and consumers of meta-analyses need to be aware of how unreliability of measurement affects heterogeneity estimates. @schmidt2015 assert that failure to control for variation in unreliability across studies results in a large overestimate of heterogeneity (p. 159), but little systematic research has been done to study the effects of unreliability on estimating heterogeneity.

@@

In classical test theory (CTT), reliability of measurement ($R_{xx}$) is defined as the proportion true score variance ($\sigma^2_T$) to observed score variance ($\sigma^2_X$) and is presumed to be constant for all participants in a sample $R_{xx} = \sigma^2_T / \sigma^2_X$. We use unreliability of measurement to refer to $1 - R_{xx}$. Imperfect reliability in measurements leads to attenuation of observed effect sizes. For product-moment correlations the observed correlation can be computed as $r_{xy} = \rho_{xy} \times \sqrt{R_{xx'}} \times \sqrt{R_{yy'}}$ where $r_{xy}$ is the observed correlation between the variables $X$ and $Y$, $\rho_{xy}$ is the true correlation (with or without sampling error) and $R_{xx'}$ and $R_{yy'}$ are the measurement reliabilities for $X$ and $Y$. As such, correcting for unreliability is straightforward for product-moment correlations, but can also be done for other effect size types [@10.1177/2515245919885611]. A corrected correlation should also have its sampling variance corrected. This is inflated by a factor equal to the square of the reliabilities, such that if $a = \sqrt{R_{xx'}} \times \sqrt{R_{yy'}}$ then the corrected sampling variance $V_\rho = V_{r} / a^2$ [e.g., @borenstein2009a, p. 343]. Reliability is usually assumed to be known [but see @10.1037/1082-989X.10.2.206; @10.1037/0021-9010.76.3.432]. Imperfect reliability in primary studies can both increase and suppress true heterogeneity.

@@

Heterogeneity is suppressed in heterogeneity estimates of uncorrected effect sizes because in absolute terms larger effect sizes are more attenuated by a given reliability. That is, although the attenuation formula $r_{xy} = \rho_{xy} \times \sqrt{R_{xx'}} \times \sqrt{R_{yy'}}$ implies that any effect size $\rho_{xy}$ will be attenuated by the same proportion given the same reliabilities $\sqrt{R_{xx'}} \times \sqrt{R_{yy'}}$, this proportion corresponds to a larger absolute value for larger effect sizes. Consequently, in the presence of heterogeneity attenuation will move larger true effect sizes further towards zero than smaller ones, decreasing heterogeneity in observed effect sizes. This suppression can be observed directly by assuming an average effect size $\theta$ with some heterogeneity $\tau^2$, infinite sample size within studies (i.e., no within-study sampling variance), and equal reliability across studies ($R = R_i = R_{xx'} = R_{yy'}$). Any observed effect size $\hat{\theta_i}$ can then be seen as a study’s true effect size $\theta_i$ multiplied by the study reliability $R$. Hence, the distribution of observed effect sizes is the distribution of true effect sizes $N(\theta, \tau^2)$ multiplied by the reliability $R$, such that $R \times N(\theta, \tau^2) = N(R \theta, R^2 \tau^2) = N(R \theta, R \tau)$. Because reliability lies between 0 – 1, under these conditions imperfect reliability implies that both the observed average effect size and the estimated heterogeneity will be less than their true value by a factor $\bar{R}$ (if heterogeneity is expressed as standard deviation rather than variance). This depiction is not entirely correct for Pearson's $r$ as it is bounded at {-1, 1}, but holds approximately if the mean effect size $\theta$ and heterogeneity $\tau^2$ are not too large. Hence, imperfect reliability in studies can be expected to suppress heterogeneity estimates whenever true heterogeneity is present. Consider an average reliability of 0.8 [the median reported reliability in psychology; @10.1177/1948550617693063]. Then, $0.8 \times N(\theta, \tau) = N(0.8\theta, 0.8 \tau)$, resulting in 20% less standard deviation among observed effect sizes than in true effect sizes. Table 2 illustrates this relationship for infinite sample size (i.e., no sampling variance) and three levels of reliability across three studies.

@@

It is difficult to predict if the total effect is a negative or positive bias in heterogeneity due to unreliability, and existing literature tends to focus on the inflation of heterogeneity [e.g., @borenstein2009a, p. 342; @card2015, p. 126; @schmidt2015; @10.1177/2515245919885611]. However, given the value of accurate heterogeneity estimates for many research outcomes in psychology and that the suppression of heterogeneity could often be 20% or more, we consider it important for meta-analysts to have insight into the bias in heterogeneity estimates that should be expected due to unreliability in a wider range of settings. The effect of unreliability depends on average effect size, true heterogeneity, mean reliability across studies, the variability in reliabilities, and sampling variance within studies. Hence, here we perform a Monte-Carlo simulation study to explore the expected bias in heterogeneity estimates due to unreliability in primary studies. In our simulation, we consider reliabilities to be known and conforming to the strict assumptions of CTT to gain a better understanding of how measurement error biases heterogeneity estimates, while acknowledging that in meta-analytic practice, heterogeneity estimation is further challenged by selective reporting, sampling error, and estimation of reliabilities, which is particularly challenging when CTT’s strict assumptions are violated.

@@

All code and data for this project are available at (OSF LINK). For our simulations and analyses we used R `r paste(R.Version()$major, R.Version()$minor, sep = ".")` [@rcoreteam2020] and we used the R package 'metafor' [@10.18637/jss.v036.i03] to perform meta-analyses and estimate heterogeneity. For parallel computing we took advantage of the R package 'parabar' [@constantin2023] and for data handling 'data.table' [@dowle2021] and 'ggplot2' [@wickham2016].

@@

We focus on correlations as our effect size type in these simulations because correcting them for unreliability is straightforward and because they are the effect size type most corrected in practice. However, unreliability attenuates all types of effect sizes and all can be corrected, either directly [standardized mean differences; @10.1177/2515245919885611] or by first transforming them into correlations. Although effect sizes can generally be transformed into each other [e.g., @borenstein2009a, p. 46], this should be done with care as it can sometimes affect conclusions [e.g., we found a violation of monotonicity in heterogeneity estimates when converting effect sizes into correlations in a previous project, @10.1037/bul0000294, supplement A].

Perhaps the most common way to meta-analyze correlations is to first transform them into Fisher z correlations [@borenstein2009a p. 41 - 43] to normalize their distribution. The Fisher $z$ transformation is the inverse hyperbolic tangent of the product moment correlation and can be computed as $z = 0.5 \times ln\left(\frac{1+r}{1-r}\right)$ and its variance approximated as $V_z = 1 / (n -3)$. There is controversy in the use of the Fisher z transformation for meta-analysis [e.g., @schmidt2015, p. 220 - 223], largely related to whether Pearson’s $r$ or Fisher’s $z$ leads to less bias in average correlations (Schulze, 2007). However, estimating heterogeneity does not seem to have been the focus of most of the literature’s discussion of the appropriateness of the Fisher z transformation [@10.1037/1082-989X.10.4.444; @10.1037/a0014697; @schmidt2015]. @10.1037/0021-9010.87.2.377 do report heterogeneity estimates but focus on the coverage of prediction intervals (‘credibility intervals’ in Hunter & Schmidt terminology) which combine the heterogeneity estimate and point estimate, and @10.1177/1094428117741966 examined results only after corrections for attenuation. Given both this inconclusive evidence and the common use of the Fisher $z$ transformation in practice, we use it as one of the effect size types in our simulations.

We also use Pearson's $r$. Pearson's $r$ has the advantage that it creates heterogeneity estimates on the same scale that are more interpretable than heterogeneity estimates on the Fisher z scale. Although Fisher $z$ correlations can be back-transformed into Pearson's $r$ for interpretation, its heterogeneity cannot. A workaround is to compute a prediction interval and convert this interval onto the product-moment scale [@10.1037/1082-989X.3.4.486]. The width of the interval can then be used as an estimate of the heterogeneity. We prefer to compute comparable heterogeneity values for Pearson's $r$ and Fisher's $z$ (see the methods section ‘Parameter values’) and report the bias in heterogeneity estimates, as we believe this is more interpretable for applied meta-analysts.

The disadvantage to using Pearson's $r$ when estimating heterogeneity are that (1) the effect size is bounded to {-1, 1} and (2) its sampling variance $\sigma^2_r = \frac{(1-\rho^2)^2}{n - 1}$ depends on the true effect size $\rho$. The bounded nature of the Pearson correlation means that large levels of heterogeneity create substantial truncation of the distribution of effect sizes if the average effect size is also large. That the sampling variance covaries with effect size leads to a small positive bias in heterogeneity estimates (supplement A). This can be solved by replacing $\rho$ with the average correlation across studies $\bar{r}$ [@schmidt2015], or using the Fisher $z$ transformation. Fisher $z$ correlations have the additional advantage that they are not bound to {-1, 1}. A disadvantage to using Fisher $z$ correlations, in addition to interpretability of heterogeneity estimates, is that they require back-transforming into product-moment correlations before adding measurement error, followed by another transformation to Fisher z for meta-analysis. Especially for large Fisher $z$ values (which are common when both average effect size and heterogeneity are large) this transformation process may introduce inaccuracies in the estimates. Given the common use of both effect size types and their mix of advantages and disadvantages we report results for both Fisher's $z$ and Pearson's $r$.

## Meta-analytic modelmet0000498

We apply two meta-analytic models to ensure that our results are not model-dependent: the @10.1037/1082-989X.3.4.486 random-effects model and the @schmidt2015 ‘bare-bones’ random-effects model. The Hedges and Vevea (HV) model uses inverse variance weight computed as $w_i = 1 / (V_i + T^2)$ [e.g., @10.1002/jrsm.12] where $V_i$ is the sampling variance of study $i$ and $T^2$ is the estimated between-studies variance. The Hunter and Schmidt (HS) model weighs studies by sample size $w_i = N_i$. The differences between these weights are relatively minor [@10.1002/jrsm.12; @schmidt2015, p. 423-426]. Although the HV model is typically applied to Fisher $z$ transformed correlations [e.g., @borenstein2009a, p. 41-43], we apply it to both transformed and untransformed correlations to enable a more direct comparison with the HS model and facilitate interpretation. The HS model typically includes corrections for unreliability or other measurement artifacts. However, we implement it without corrections [‘bare-bones’, @schmidt2015] because this allows us to focus on the bias in heterogeneity estimates rather than the performance of corrections [e.g., see @10.1037/met0000498 for how well they perform in terms of mean estimates].
